A Deep Learning Approach for Gait Event Detection from a Single Shank-Worn IMU: Validation in Healthy and Neurological Cohorts

### 2022-03-18 paper/MDPI_template/article.tex
Updated paper

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, adolescents, aerospace, agriculture, agriengineering, agronomy, ai, algorithms, allergies, alloys, analytica, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coasts, coatings, colloids, colorants, commodities, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryptography, crystals, csmf, ctn, curroncol, currophthalmol, cyber, dairy, data, dentistry, dermato, dermatopathology, designs, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, ent, entomology, entropy, environments, environsciproc, epidemiologia, epigenomes, est, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, foundations, fractalfract, fuels, futureinternet, futureparasites, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, hazardousmatters, healthcare, hearts, hemato, heritage, highthroughput, histories, horticulturae, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijtm, ijtpp, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jzbg, kidney, kidneydial, knowledge, land, languages, laws, life, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, microplastics, minerals, mining, modelling, molbank, molecules, mps, msf, mti, muscles, nanoenergyadv, nanomanufacturing, nanomaterials, ncrna, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, nri, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacoepidemiology, pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, pollutants, polymers, polysaccharides, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, publications, quantumrep, quaternary, qubs, radiation, reactions, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, ruminants, safety, sci, scipharm, seeds, sensors, separations, sexes, signals, sinusitis, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, sports, standards, stats, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, taxonomy, technologies, telecom, test, textiles, thalassrep, thermo, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, viruses, vision, waste, water, wem, wevj, wind, women, world, youth, zoonoticdis

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by: 
% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}, where the beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or stereophotogrammetry systems based on reflective markers using a multi-camera setup \cite{Chiari2005}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is more and more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

@@

In this study, we further these works by validating this deep learning approach for detecting gait events in a heterogeneous cohort of healthy and neurologically diseased adults considering a single IMU setup, worn on the lower leg.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

\subsection{Data Collection}
Gait analyses were performed in the Universit\"{a}tsklinikum Schleswig-Holstein (UKSH) campus Kiel, Germany. The study \cite{Warmerdam2021} was approved by the ethical committee of the medical faculty at the UKSH (no: D438/18). In total, data from 160 participants were included for the current analysis, including data from young adults (YA; age: 18 - 60 years), older adults (OA; age: $>$60 years), people with Parkinson's Disease (PD; according to the UK Brain Bank criteria \cite{Gibb1988}), people with a recent ($<$4 weeks) symptomatic stroke (stroke), people with multiple sclerosis (MS; according to the McDonalds criteria \cite{Thompson2018}), people with chronic low back pain (cLBP), and people with other diagnoses that were assumed not to affect mobility (Table \ref{tab:demographics_data}). Inclusion criteria were an age of 18 years or older, and the ability to walk independently without a walking aid. Participants were excluded from the study with a Montreal Cognitive Assessment \cite{Nasreddine2005} score $<$15 and other movement disorders that affected mobility, as noticed by the clinical assessor.

@@

@@

For the current analysis data from four IMUs (Noraxon USA Inc., myoMOTION, Scottsdale, AZ, USA) were considered, namely those that were attached laterally above the left and right ankle joint and those attached proximally at the left and right shank. IMUs were secured to participants using elastic bands with a special hold for the IMU. Furthermore, reflective markers were attached on top of the usual foot wear at the heel and toe of both feet. Marker data were recorded using a twelve-camera stereophotogrammetry system (Qualisys AB, G\"{o}teborg, Sweden) at a sampling frequency of 200 Hz. IMU data were recorded at the same sampling frequency, and both systems were synchronized using TTL signal \cite{Warmerdam2021}. For some recordings, the sampling frequency of the IMUs was erroneously set at 100 Hz, and therefore data from these trials were upsampled before further analysis.

\subsection{Data Pre-processing \label{subsec:data_preprocessing}}
\subsubsection{Reference Values}
The raw reflective marker data were first used to determine the start and end of each trial. The start of the trial was defined as the time for which the first toe marker (left or right) crossed the \emph{virtual} starting line. The end was defined as the time for which the last heel marker crossed the end line. For both marker and IMU systems, data were then cropped from start to the end of the trial.

Any remaining gaps in the marker data were filled by interpolation making use of inter-correlations between markers \cite{Federolf2013,Gloersen2016}. The data were then low-pass filtered using a double pass 6th order Butterworth filter with a cut-off frequency of 20 Hz \cite{Kormylo1974,Racz2021}. The filtered data were differentiated to get velocity signals, and timings of ICs and FCs were determined from local maxima and minima in the heel and toe vertical velocity signals \cite{Pijnappels2001,OConnor2007}. Like in \cite{Carcreff2018,Romijnders2021}, all identified ICs and FCs were manually checked using Qualisys Track Manager 2018.1 software (QTM; Qualisys AB, Göteborg, Sweden).

\subsubsection{IMU Data}
For the deep learning model, the data from randomly selected one-third of participants were used as a "training set", another one-third was used as a "validation set", and the rest was used as "test set", where randomization was stratified by both group and gender. Accelerometer and gyroscope data from the four IMUs were normalized by subtracting the channel-wise mean, and dividing by the channel-wise standard deviation. Then, for the training and validation set, the time series data were partitioned into equal length time windows of 400 samples, with an overlap of 50\% between successive windows.

\subsection{Model}
\subsubsection{Notation}
Data from the IMUs' 3D accelerometer and 3D gyroscope were used. A signal was considered a sequence of real-valued numbers: $\mathbf{x} = \begin{array}{ccccc}
	\cdots & x[n-1] & x[n] & x[n+1] & \cdots
\end{array}$, with $n = 1, \cdots, N$ the discrete-time sample index. For any given channel, e.g., acceleration in $X$-direction, the signal was denoted in vector notation as: $\mathbf{x}_{d} = \begin{bmatrix}
x_{d}[1] & x_{d}[2] & \cdots & x_{d}[N]
\end{bmatrix}^{\mathrm{T}}$, where $d$ referred to the $d$-th channel.

Data from all channels of a single IMU were collected in a $ N \times D$ data matrix, like:
\begin{equation}
	\mathbf{X} = \begin{bmatrix}
		\vdots & \vdots &  & \vdots\\
		\mathbf{x}_{1} & \mathbf{x}_{2} & \cdots & \mathbf{x}_{D}\\
		\vdots & \vdots &  & \vdots
	\end{bmatrix} = \begin{bmatrix}
		x_{1}[1] & x_{2}[1] & & x_{D}[1]\\
		x_{1}[2] & x_{2}[2] & & x_{D}[2]\\
		\vdots & \vdots & \cdots & \vdots\\
		x_{1}[N] & x_{2}[1] & & x_{D}[N]\end{bmatrix}, \quad \mathbf{X} \in \mathbb{R}^{N \times D}
	\label{eqn:train_data_single_instance}
\end{equation}

The corresponding labels (or targets) were then denoted as:
\begin{equation}
	\mathbf{y}_{\mathrm{IC}} = \begin{bmatrix}
		y_{1}[1]\\
		y_{1}[2]\\
		\vdots\\
		y_{1}[N]
	\end{bmatrix}, \quad \mathbf{y}_{\mathrm{FC}} = \begin{bmatrix}
		y_{2}[1]\\
		y_{2}[2]\\
		\vdots\\
		y_{2}[N]
	\end{bmatrix}, \quad y_{i}[n] \in [0, 1]
	\label{eqn:train_labels_single_instace}
\end{equation}

\subsubsection{Model Architecture}
The basic architecture for the deep learning model was a temporal convolutional network (TCN) \cite{Bai2018,Remy2020,Filtjens2020}. The TCN consisted of repeating blocks of dilated convolutions (CONV) \cite{YuKoltun2016,Bai2018} that were followed by batch normalization (BN) \cite{Ioffe2015}, rectified linear unit (ReLU) activation, and dropout (DropOut) \cite{Srivastava2014}. For each given dilation factor, the sequence of layers CONV-BN-ReLU-DropOut was repeated twice \cite{Remy2020}. Dilation factors were given as a sequence of increasing powers of 2, e.g., $\left\{1, 2, 4, 8\right\}$ \cite{VanDenOord2016,YuKoltun2016,Bai2018}. These repeating blocks were followed by a fully-connected (dense) layer with sigmoid activation, and outputs were predicted separately for ICs and FCs. The mean squared error (MSE) was used as loss function, and a gradient descent-based optimization algorithm with adaptive moment (Adam) optimizer was used to iteratively learn the weights \cite{Kingma2014,Schmidt2021}.

@@

\subsubsection{Time agreement}
For all correctly detected gait events (TP, Section \ref{subsubsec:overall_detection_performance}), the time error between the annotated and detect gait event was defined as:
\begin{equation}
	\textrm{time error} = t_{\textrm{ref}} - t_{\textrm{pred}}
	\label{eqn:time_agreement}
\end{equation}
with $t_{\mathrm{ref}}$ the gait event time from the marker-based annotations, and $t_{\mathrm{pred}}$ the gait event time from the model predictions. As a robust measure for the average time error and its spread, the median time error and the inter-quartile range (IQR) were reported \cite{OpenIntro2019}.

\subsubsection{Gait parameters}
For those trials for which all gait events were detected, and no spurious events were detected, the following stride parameters were calculated:
\begin{equation}
	\textrm{stride time} = t_{\textrm{IC}_{k}} - t_{\textrm{IC}_{k-1}}
	\label{eqn:stride_time}
\end{equation}
\begin{equation}
	\textrm{stance time} = t_{\textrm{FC}_{k}} - t_{\textrm{IC}_{k-1}}
	\label{eqn:stance_time}
\end{equation}
\begin{equation}
	\textrm{swing time} = t_{\textrm{IC}_{k}} - t_{\textrm{FC}_{k}}
	\label{eqn:swing_time}
\end{equation}

@@

For both ICs and FCs, recall is high for each of the tracked points (i.e, $\ge 92\%$), and so is precision (i.e, $\ge 97\%$). Differences between the tracked points are small, i.e. the minimum recall is $92\%$ and the maximum recall is $97\%$, and the minimum precision is $97\%$ and the maximum precision is $99\%$.

\subsection{Time error}
\begin{figure}[H]
	\includegraphics[width=10.5 cm]{fig/box_plots_tracked_points_with_outliers.png}
	\caption{Time errors for initial (left) and final (right) contacts detection, for each of the different tracked points.\label{fig:time_error_box_plots}}
\end{figure}

Time errors for ICs and FCs are visually depicted using boxplots for each tracked point. Data were not normally distributed, thus Wilcoxon signed-rank tests were used to evaluate whether the median time error was zero.
\begin{table}[H] 
	\caption{Time errors for the correctly detected gait events. \label{tab:time_errors}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{r|CCC|CCC}
		\toprule
		 & \multicolumn{3}{c}{\textbf{Initial contacts}} & \multicolumn{3}{c}{\textbf{Final contacts}}\\
		\textbf{Tracked point} & \textbf{median} & \textbf{IQR} & (\textbf{Q25}, \textbf{Q75}) & \textbf{median} & \textbf{IQR} & (\textbf{Q25}, \textbf{Q75})\\
		 & ms & ms &  (ms, ms) & ms & ms & (ms, ms)\\
		\midrule
		left ankle & 0 & 20 & (-10, 10) & 0 & 10 & (-5, 5)\\
		right ankle & 0 & 20 & (-15, 5) & -5 & 15 & (-10, 5)\\
		left shank & -5 & 20 & (-10, 10) & -5 & 20 & (-15, 5)\\
		right shank & -2.5 & 20 & (-15, 5) & -5 & 20 & (-15, 5)\\
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{IQR: inter-quartile range, Q25: 25th percentile, Q75: 75th percentile.}}
\end{table}

@@

\begin{figure}[H]
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
		\includegraphics[width=13.5cm]{fig/bland_altman_plots_stride_params.png}
	\end{adjustwidth}
	\caption{The agreement of extracted gait parameters between the sensor-based and marker-based methods.\label{fig:gait_parameters_bland_altman_plots}}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided. The following statements should be used ``Conceptualization, X.X. and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y. and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing---original draft preparation, X.X.; writing---review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y. All authors have read and agreed to the published version of the manuscript.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation. Authorship must be limited to those who have contributed substantially to the work~reported.}

\funding{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''. Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}

\institutionalreview{In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study. You might choose to exclude this statement if the study did not require ethical approval. Please note that the Editorial Office might ask you for further information. Please add “The study was conducted in accordance with the Declaration of Helsinki, and approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving humans. OR “The animal study protocol was approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving animals. OR “Ethical review and approval were waived for this study due to REASON (please provide a detailed justification).” OR “Not applicable” for studies not involving humans or animals.}

>>>>

%--------------------
% Class Options:
%--------------------
%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}, where the beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or passive, retro-reflective marker-based optoelectronic motion capture systems (henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is more and more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

@@

In this study, we further these works by validating this deep learning approach for detecting gait events in a heterogeneous cohort of healthy and neurologically diseased adults considering a single IMU setup, worn on the lower leg.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
\subsection{Data Collection}
Gait analyses were performed in the Universit\"{a}tsklinikum Schleswig-Holstein (UKSH) campus Kiel, Germany. The study \cite{Warmerdam2021} was approved by the ethical committee of the medical faculty at the UKSH (no: D438/18). In total, data from 160 participants were included for the current analysis, including data from young adults (YA; age: 18 - 60 years), older adults (OA; age: $>$60 years), people with Parkinson's Disease (PD; according to the UK Brain Bank criteria \cite{Gibb1988}), people with a recent ($<$4 weeks) symptomatic stroke (stroke), people with multiple sclerosis (MS; according to the McDonalds criteria \cite{Thompson2018}), people with chronic low back pain (cLBP), and people with other diagnoses that were assumed not to affect mobility (Table \ref{tab:demographics_data}). Inclusion criteria were an age of 18 years or older, and the ability to walk independently without a walking aid. Participants were excluded from the study with a Montreal Cognitive Assessment \cite{Nasreddine2005} score $<$15 and other movement disorders that affected mobility, as noticed by the clinical assessor.

@@

\begin{figure}[H]
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
		\includegraphics[width=13.5cm]{fig/methods_overall_workflow.pdf}
	\end{adjustwidth}
	\caption{Schematic depiction$^{1}$ of the current study. Study participants wore IMUs on the ankle and shanks, and reflective markers were adhered on the heel and toe of usual footwear (shown on the left). Marker data were used to obtain reference values for the timings of initial and final contacts (top), where accelerometer and gyroscope data from each tracked point were inputted to a neural network that predicted timings of the same initial and final contacts (bottom). \newline \noindent{\footnotesize{\textsuperscript{1} Picture from: \url{https://www.vecteezy.com/free-vector/man-walking }.}} \label{fig:methods_marker_vs_imu}}
\end{figure}

@@

For the current analysis data from four IMUs (Noraxon USA Inc., myoMOTION, Scottsdale, AZ, USA) were considered, namely those that were attached laterally above the left and right ankle joint and those attached proximally at the left and right shank. IMUs were secured to participants using elastic bands with a special hold for the IMU. Furthermore, reflective markers were attached on top of the usual foot wear at the heel and toe of both feet (Figure~\ref{fig:methods_marker_vs_imu}). Marker data were recorded using a twelve-camera optoelectronic motion capture system (Qualisys AB, G\"{o}teborg, Sweden) at a sampling frequency of 200 Hz. IMU data were recorded at the same sampling frequency, and both systems were synchronized using a TTL signal \cite{Warmerdam2021}. For some recordings, the sampling frequency of the IMUs was erroneously set at 100 Hz, and therefore data from these trials were upsampled before further analysis.

\subsection{Data Pre-processing \label{subsec:data_preprocessing}}
\subsubsection{Marker Data}
The reflective marker data were first used to determine start and end of each trial. The start of the trial was defined as the time for which the first toe marker (left or right) crossed the \emph{virtual} starting line. The end was defined as the time for which the last heel marker crossed the end line. For both marker and IMU systems, data were then cropped from start to the end of the trial.

Any remaining gaps in the marker data were filled by interpolation making use of inter-correlations between markers \cite{Federolf2013,Gloersen2016}. The data were then low-pass filtered using a double pass 6th order Butterworth filter with a cut-off frequency of 20 Hz \cite{Kormylo1974,Racz2021}. The filtered data were differentiated to get velocity signals, and timings of ICs and FCs were determined from local maxima and minima in the heel and toe vertical velocity signals \cite{Pijnappels2001,OConnor2007}. Like in \cite{Carcreff2018,Romijnders2021}, all identified ICs and FCs were manually checked, and corrected if necessary, using Qualisys Track Manager 2018.1 software (Qualisys AB, Göteborg, Sweden). The resulting annotated ICs and FCs were considered the \emph{true} events (also \emph{labels} or \emph{targets}), and were used as reference timings to derive stride-specific gait parameters.

\subsubsection{IMU Data}
The idea behind the deep learning approach was that a \emph{model} was trained to predict the likelihood of an IC and FC, given accelerometer and gyroscope data from a single IMU. In other words, the model was iteratively trained to learn a mapping $h_{\mathbf{W}}(\mathbf{X}): \mathbf{X} \rightarrow \mathbf{y}$, where $h_{\mathbf{W}}$ was also referred to as the \emph{hypothesis} parameterized by the weights, collectively denoted by $\mathbf{W}$, $\mathbf{X}$ was the matrix of IMU data, and $\mathbf{y}$ was the vector of annotated events.

Here, the matrix $\mathbf{X}$ consisted of the IMU data, arranged as:
\begin{equation}
	\mathbf{X} = \begin{bmatrix}
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}\\
		\mathbf{x}_{1} & \mathbf{x}_{2} & \cdots & \mathbf{x}_{D}\\
		\rule[0ex]{0.5pt}{2.5ex} & \rule[0ex]{0.5pt}{2.5ex} &  & \rule[0ex]{0.5pt}{2.5ex}
	\end{bmatrix} = \begin{bmatrix}
		x_{1}[1] & x_{2}[1] & & x_{D}[1]\\
		x_{1}[2] & x_{2}[2] & & x_{D}[2]\\
		\vdots & \vdots & \cdots & \vdots\\
		x_{1}[N] & x_{2}[1] & & x_{D}[N]\end{bmatrix}, \quad \mathbf{X} \in \mathbb{R}^{N \times D}
	\label{eqn:train_data_single_instance}
\end{equation}
where $\mathbf{x}_{d} = \begin{bmatrix}x_{d}[1] & x_{d}[2] & \cdots & x_{d}[n] & \cdots & x_{d}[N]\end{bmatrix}^{\mathrm{T}}$, with $d = 1, \cdots, D$ referred to the $d$-th channel, and $n = 1, \cdots, N$ referred to the $n$-th sample (or discrete time step).

Accordingly, the vector $\mathbf{y}$ consisted of the annotated events, arranged as:
\begin{equation}
	\mathbf{y}_{\mathrm{IC}} = \begin{bmatrix}
		y_{1}[1]\\
		y_{1}[2]\\
		\vdots\\
		y_{1}[N]
	\end{bmatrix}, \quad \mathbf{y}_{\mathrm{FC}} = \begin{bmatrix}
		y_{2}[1]\\
		y_{2}[2]\\
		\vdots\\
		y_{2}[N]
	\end{bmatrix}, \quad y_{i}[n] \in [0, 1]
	\label{eqn:train_labels_single_instace}
\end{equation}

The data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}).

Accelerometer and gyroscope data were normalized by subtracting the channel-wise mean, and dividing by the channel-wise standard deviation. Then, for the training and validation data set, the data were partitioned into equal length time windows \cite{Filtjens2020} of 400~samples, with an overlap of 50\% between successive windows (corresponding to 2 s windows, and 1 s overlap, respectively).

\subsection{Model}
\subsubsection{Model Architecture}
The basic architecture for the deep learning model was a temporal convolutional network (TCN) \cite{Bai2018,Remy2020,Filtjens2020}. The TCN consisted of repeating blocks of dilated convolutions (CONV) \cite{YuKoltun2016,Bai2018} that were followed by batch normalization (BN) \cite{Ioffe2015}, rectified linear unit (ReLU) activation, and dropout (DropOut) \cite{Srivastava2014}. For each given dilation factor, the sequence of layers CONV-BN-ReLU-DropOut was repeated twice \cite{Remy2020}. Dilation factors were given as a sequence of increasing powers of 2, e.g., $\left\{1, 2, 4, 8\right\}$ \cite{VanDenOord2016,YuKoltun2016,Bai2018}. These repeating blocks were followed by a fully-connected (dense) layer with sigmoid activation, and outputs were predicted separately for ICs and FCs. The mean squared error (MSE) was used as loss function, and a gradient descent-based optimization algorithm with adaptive moment (Adam) optimizer was used to iteratively learn the weights \cite{Kingma2014,Schmidt2021}.

@@

\subsubsection{Time Error}
For all correctly detected gait events (TP, Section \ref{subsubsec:overall_detection_performance}), the time error between the annotated and detect gait event was defined as:
\begin{equation}
	\textrm{time error} = t_{\textrm{ref}} - t_{\textrm{pred}}
	\label{eqn:time_agreement}
\end{equation}
with $t_{\mathrm{ref}}$ the gait event time from the marker-based annotations, and $t_{\mathrm{pred}}$ the gait event time from the model predictions. As a robust measure for the average time error and its spread, the median time error and the inter-quartile range (IQR) were reported \cite{OpenIntro2019}.

\subsubsection{Stride Parameters}
For those trials for which all gait events were detected, and no spurious events were detected, the stride time, stance time and swing were calculated. Stride was the time between two successive ICs of the same foot. Stance time was the time between FC and the preceding IC of the same foot. Swing time was the time between the IC following the last FC of the same foot.

@@

For both ICs and FCs, recall is high for each of the tracked points (i.e., $\ge 92\%$), and so is precision (i.e., $\ge 97\%$). Differences between the tracked points are small, i.e. the minimum recall is $92\%$ and the maximum recall is $97\%$, and the minimum precision is $97\%$ and the maximum precision is $99\%$.

\subsection{Time Error}
\begin{figure}[H]
	\centering
	\includegraphics[width=13.5 cm]{fig/box_plots_tracked_points_with_outliers.png}
	\caption{Time errors for initial (left) and final (right) contacts detection, for each of the different tracked points.\label{fig:time_error_box_plots}}
\end{figure}

Time errors for ICs and FCs are visually depicted using boxplots for each tracked point. Data were not normally distributed, thus Wilcoxon signed-rank tests were used to evaluate whether the median time error was zero.
\begin{table}[H] 
	\caption{Time errors for the correctly detected gait events. \label{tab:time_errors}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{r|CC|CC}
		\toprule
		 & \multicolumn{2}{c}{\textbf{Initial contacts}} & \multicolumn{2}{c}{\textbf{Final contacts}}\\
		\textbf{Tracked point} & \textbf{median} & \textbf{IQR} & \textbf{median} & \textbf{IQR}\\
		 & s & s & s & s \\
		\midrule
		left ankle & 0.000 & 0.020 & 0.000 & 0.010\\
		right ankle & 0.000 & 0.020 & -0.005 & 0.015\\
		left shank & -0.005 & 0.020 & -0.005 & 0.020\\
		right shank & -0.003 & 0.020 & -0.005 & 0.020\\
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{IQR: inter-quartile range.}}
\end{table}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=13.5cm]{fig/bland_altman_plots_stride_params.png}
	\caption{The agreement of extracted gait parameters between the sensor-based and marker-based methods.\label{fig:gait_parameters_bland_altman_plots}}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
In this study we have validated a deep learning-based approach to detect gait events and subsequently extract clinically relevant stride parameters from a single inertial measurement unit. Performance analysis showed an excellent detection rate, and low time errors in both event detection and stride parameter calculation for different walking speeds and across both healthy and neurological cohorts. Our next step is to validate these methods under more challenging conditions, that involve real-life walking sequences.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{Conceptualization, R.R., G.S., and W.M.; methodology, R.R., G.S.; software, R.R.; validation, R.R.,; formal analysis, R.R.; investigation, R.R., G.S., W.M.; resources, C.H, W.M..; data curation, E.W., R.R.,; writing---original draft preparation, R.R.; writing---review and editing, E.W., C.H., G.S., W.M.; visualization, R.R.; supervision, G.S., W.M.; project administration, R.R., E.W., C.H., G.S., W.M.; funding acquisition, W.M. All authors have read and agreed to the published version of the manuscript.}

\funding{This research received funding from the DFG Open Access fund.}

\institutionalreview{The study was conducted in accordance with the Declaration of Helsinki, and approved by the Ethics Committee of the medical faculty of the Christian-Albrechts-Universit\"{a}t zu Kiel (D438/18, approved on 8 May 2018).}

\informedconsent{Informed consent was obtained from all subjects involved in the study.}

\dataavailability{Data from the first 10 participant are available online at \url{https://github.com/neurogeriatricskiel/Validation-dataset}. Additionally, we are preparing the open-source release of all data from the healthy younger and older adults. Data from ptient groups }

\acknowledgments{The authors sincerely thank all people that were involved in the data collection, from study participants to students who aided and assisted during measurements and subsequent marker labeling. The authors appreciate the off- and on-topic discussions with Julius Welzel, and his positive contributions to the organization of the research data. The authors give a thumbs up to Johannes Hoffmann for his tips and tricks regarding the \LaTeX typesetting.}

\conflictsofinterest{The authors declare no conflict of interest.}

%% Optional
% \sampleavailability{Samples of the compounds ... are available from the authors.}
